---
title: "Relationship between mpg and other variables"
author: "Helen"
date: "30 November 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, cache = TRUE)
library(datasets)
library(ggplot2)
library(broom)
library(dplyr)
library(tidyr)
```

## Executive Summary

This report investigates the relationship between mpg and other variables in the mtcars dataset. In particular, the report aims to answer two questions:

1. Is an automatic or manual transmission better for mpg?
2. Quantify the mpg difference between automatic and manual transmissions.

## The data

The dataset mtcars consists of 11 variables recorded for 32 different cars. The first six rows of the dataset are shown below.

```{r data sample}
head(mtcars)
```

This report is concerned mainly with columns 1 and 9, which are mpg (miles per gallon) and transmission type (0 = automatic and 1 = manual).

An initial exploratory plot is shown below.

```{r initial mpg am plot}
ggplot(mtcars, aes(x = as.factor(am), y = mpg)) + geom_point()
```

It can be seen from the plot that automatic transmissions (am = 0) tend to have lower mpgs but there is some overlap. The spread of the points can be seen from the boxplot below.

```{r mpg am boxplot}
ggplot(mtcars, aes(x = as.factor(am), y = mpg))+ geom_boxplot()
```

## Fitting models

The first model fitted was a linear model with a single variable. 

```{r lm1}
fit1 <- lm(mpg ~ factor(am), mtcars)
summary(fit1)
```

This model indicates that automatic transmission cars have a mean mpg of 17.147 and manual transmission cars have a mean mpg of (17.147 + 7.245 =) 24.392. The p-value for this difference is 0.000285 (i.e. around 0.3%) so is significant at the 5% level. 

The second model fitted uses weight (wt) as another variable.

```{r lm2}
fit2 <- lm(mpg ~ factor(am) + wt, mtcars)
summary(fit2)
```

This model indicates that, when weight is held constant, the difference in mpg between automatic and manual transmission cars is negligible. The coefficient has changed from +7.245 to -0.02362, meaning that for a specific car weight, the expected mpg for a manual car is 0.02362 less than for an automatic car. It should be noted that this difference is not statistically significant at all. 

## Understanding the effect of weight

This result can be better understood by considering the graph below.

```{r mpg graph am wt}
ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point() + facet_grid(rows = mtcars$am)
```

The graph above shows that automatic cars tend to be heavier than manual cars, and this difference in weight almost entirely explains the differences in mpg between automatic and manual cars.

The graph below shows that when automatic and manual cars are considered separately both sets are data are strongly influenced by extreme points.

```{r leverage}
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(am))) + geom_point() + geom_smooth(method = "lm")
```

The size of the effect of the outer points can be calculated if separate models are fitted for automatic and manual transmissions, as in the graph above.

```{r separate models}
fitauto <- lm(mpg ~ wt, mtcars[mtcars$am == 0, ])
resultsauto <- as_tibble(data.frame(name = rownames(mtcars[mtcars$am == 0, ]), mpg = mtcars[mtcars$am == 0, 1], wt = mtcars[mtcars$am == 0, 6], hv = hatvalues(fitauto),  dffits = dffits(fitauto))) %>% mutate(av_hv = hv / mean(hv))
fitmanual <- lm(mpg ~ wt, mtcars[mtcars$am == 1, ])
resultsmanual <- as_tibble(data.frame(name = rownames(mtcars[mtcars$am == 1, ]), mpg = mtcars[mtcars$am == 1, 1], wt = mtcars[mtcars$am == 1, 6], hv = hatvalues(fitmanual), dffits = dffits(fitmanual))) %>% mutate(av_hv = hv/mean(hv))
resultsauto
resultsmanual
```

Large hatvalues and large dffits indicate large influence. It can be seen that 12, 13, 14 and 15 for automatic have at least one of these values being large.

Fitting a model across both transmission types together removes the influential effect of some of these points.

```{r results together}
ggplot(mtcars, aes(x = mpg, y = wt, col = factor(am))) + geom_point() + geom_smooth(method = "lm", aes(group = 1))
fit3 <- lm(mpg ~ wt, mtcars)
resultsall <- as_tibble(data.frame(name = rownames(mtcars), mpg = mtcars$mpg, wt = mtcars$wt, hv = hatvalues(fit3), dffits(fit3)))
summary(resultsall$hv)
```

The previously influential points around the middle are less influential now. Likely to be due to sparse data in this area when considering automatic and manual separately.

## Check for fit

The residuals are plotted in the graph below to check for fit.

```{r lm2 residuals}
ggplot(fit2, aes(x = .fitted, y = .resid)) + geom_point()
```

There is a slight trend in the residuals to be larger at the small and large x values. This suggests that model fit is not ideal.

The ANOVA analysis below shows that the inclusion of transmission type does not appear to add any additional significant information beyond that included by weight.

```{r anova}
anova(lm(mpg ~ wt, mtcars), lm(mpg ~ wt + factor(am), mtcars))
```

The difference between the models as a result of adding in transmission type is negligible, and nowhere near a 5% p-value.

